Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
method for parallel

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
\newcommand{\priorityq}[1]{\mbox{PriorityQueue}({#1})}
\newcommand{\listnew}[1]{\mbox{List}(#1)}
\newcommand{\pqpush}[2]{#1.\mbox{push}(#2)}
\newcommand{\pqpop}[1]{#1.\mbox{pop}()}
\newcommand{\listappend}[2]{#1.\mbox{append}(#2)}
\newcommand{\ancsegment}[1]{\mbox{Segment}(#1)}
\newcommand{\taddrow}[2]{#1.\mbox{addrow}(#2)}
\newcommand{\attrparent}[1]{#1.\mbox{parent}}
\newcommand{\attrchild}[1]{#1.\mbox{child}}
\newcommand{\attrleft}[1]{#1.\mbox{left}}
\newcommand{\attrright}[1]{#1.\mbox{right}}
\newcommand{\attrid}[1]{#1.\mbox{id}}
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one to two orders of magnitude.
    In addition to speed, the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored in $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation (and its' history)
    rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange
    of tree sequences between modular program components,
    and show how to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
\end{abstract}

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
and, hence, genome sequence, as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g.,
\citet{gog2014theory}).
Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).
Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
These made use of the GNU Scientific Library \citep[version 1.16][]{galassi2018scientific},
pybind11 \citep[version 2.2.1][]{pybind11},
and GCC (version 4.8.5).
We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
the Wellcome Trust (grant 100956/Z/13/Z to Gil McVean);
and XXX.

@@

\algstep{S0.}{Initialize.}{
    Set $\Al_u \leftarrow \listnew{}$ for $0 \leq u < |\Nt_I|$. Then,
    for each $u$ in $S$, set $v \leftarrow \taddrow{\Nt_O}{\Nt_I[u]}$ and
    call $\listappend{\Al[u]}{\ancsegment{0, L, v}}$. Then set $Q \leftarrow \priorityq{}$
    and $u \leftarrow 0$.
}

\algstep{S1.}{Select parent edges.}{
    Set $E \leftarrow \{e: \attrparent{e} = u \mbox{ and } e \in \Et_I\}$.
    Set $v \leftarrow -1$.
}

\algstep{S2.}{Insert edge/ancestry intersections.}{
    For each $e\in E$ and each $x \in \Al[\attrchild{e}]$,
    if $\attrright{x} > \attrleft{e}$ and $\attrright{e} > \attrleft{x}$,
    set $\ell \leftarrow \max(\attrleft{x}, \attrright{e})$,
    $r \leftarrow \min(\attrright{x}, \attrright{e})$,
    $y \leftarrow \ancsegment{\ell, r, \attrid{x}}$ and call
    $\pqpush{Q}{y}$.
}

\algstep{S3.}{Find segments with minimum left coordinate.}{
    Set $\ell \leftarrow \attrleft{Q[0]}$, $r \leftarrow L$ and $X \leftarrow \mbox{List}()$.
    Then, while $|Q| > 0$ and $\attrleft{Q[0]} = \ell$, set $x \leftarrow \pqpop{Q}$,
    $r \leftarrow \min(r, \attrright{x})$ and call $\listappend{X}{x}$.
    Afterwards, if $|Q| > 0$, set $r \leftarrow \min(r, \attrleft{Q[0]})$. If
    $|X| > 1$ go to \algref{S5}.
}

\algstep{S4.}{No overlapping segments.}{
    Set $x \leftarrow X[0]$ and $\alpha \leftarrow x$. If $|Q| > 0$ and
    $\attrleft{Q[0]} < \attrright{x}$, set $\alpha \leftarrow
    \ancsegment{\attrleft{x}, \attrright{Q[0]}, \attrid{x}}$,
    $\attrleft{x} \leftarrow \attrleft{Q[0]}$ and call $\pqpush{Q}{x}$.
    Go to \algref{S7}.
}

\algstep{S5.}{New output node.}{
    If $v = -1$, set $v \leftarrow \taddrow{\Nt_0}{\Nt_I[u]}$.
}

\algstep{S6.}{Record edges.}{
    Set $\alpha \leftarrow \ancsegment{\ell, r, v}$. Then for each $x \in X$
    call $\taddrow{\Et_O}{l, r, v, \attrid{x}}$ and, if $\attrright{x} > r$,
    set $\attrleft{x} \leftarrow r$ and call $\pqpush{Q}{x}$.
}

\algstep{S7.}{Left coordinate loop.}{
    Call $\listappend{\Al}{\alpha}$. If $|Q| > 0$ return to \algref{S3}.
}

\algstep{S8.}{Parent loop.}{
    Set $u \leftarrow u + 1$. If $u < |\Nt_I|$, go to \algref{S1}.
}

@@

@@

>>>>

\include{macros}
\newcommand{\simupop}{\texttt{simuPOP}}
\newcommand{\fwdpp}{\texttt{fwdpp}}
\newcommand{\fwdpy}{\texttt{fwdpy11}}
\newcommand{\justc}{\texttt{C}}
\newcommand{\cpp}{\texttt{C++}}
\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\ftprime}{\texttt{ftprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}
\newcommand{\setdiff}{\smallsetminus}
\newcommand{\Nt}{\mathcal{N}}  % node table
\newcommand{\Et}{\mathcal{E}}  % edge table
\newcommand{\St}{\mathcal{S}}  % site table
\newcommand{\Mt}{\mathcal{M}}  % mutation table
\newcommand{\Al}{\mathcal{A}}  % list of ancestors
% mbox to avoid hyphenating
\newcommand{\taddnode}[1]{\mbox{\texttt{addnode}}(#1)}
\newcommand{\taddedge}[1]{\mbox{\texttt{addedge}}(#1)}
\newcommand{\tsimplify}[1]{\mbox{\texttt{simplify}}(#1)}
\newcommand{\tsort}[1]{\mbox{\texttt{sort}}(#1)}
\newcommand{\tnodetable}[1]{\mbox{\texttt{NodeTable}}(#1)}
\newcommand{\tedgetable}[1]{\mbox{\texttt{EdgeTable}}(#1)}
\DeclareMathOperator{\len}{length}
\DeclareMathOperator{\add}{add} % for adding to tables
\newcommand{\lleft}{\mathrm{left}}  % left label
\newcommand{\lright}{\mathrm{right}}
\newcommand{\lid}{\mathrm{id}}

@@

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one to orders of magnitude.
    In addition to speed, the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored in $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation (and its' history)
    rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange 
    of tree sequences between modular program components,
    and show how to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
\end{abstract}

@@

In this paper, we describe a storage method for \emph{succinct tree sequences} 
and, hence, genome sequence, as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
while still allowing efficient retrieval of information.
Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

We implemented simulations and the connection to \msprime{} in \cpp{}, using \fwdpp{} library functions and
interface code using a continuum-sites model for both mutation
and recombination. Simulations were run using \fwdpy{} (version 0.13.a0), a Python package based
on \fwdpp{} (version 0.5.7).  The majority of results are presented based on a single-threaded implementation.  However,
we also implemented a parallelized version using Python's \texttt{queue.Queue} to run the simplification step in a separate Python
thread. Our implementation allows a maxmimum of four simplification intervals to be in the queue at once. This
parallelized version also performed fitness calculation in parallel using two threads of execution in C++.

Code for all simulations and figures is available at \url{https://github.com/petrelharp/ftprime_ms}.
These made use of the GNU Scientific Library \citep[version 1.16][]{galassi2018scientific},
pybind11 \citep[version 2.2.1][]{pybind11},
and GCC (version 4.8.5).
We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 GHz Intel E5-2650 CPU with
hyperthreading enabled.
We ran one simulation at a time and the machine was under minimal load otherwise.
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
%% Our simulation scripts separately recorded the time spent in various steps of each run.

%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
Thanks to Jared Galloway, Brad Shaffer, and Evan McCartney--Melstad for useful discussions.
Work on this project was supported by funding from 
the Sloan Foundation and the NSF (under DBI-1262645) to PR;
and XXX.

@@

\algstep{S0.}{Initialize.}{
    Set $\Al[j] \leftarrow (0, L, j)$
    and $\Nt_O[j] \leftarrow \Nt_I[S[j]]$ for $0 \le j < \len(S)$.
    Set the current input parent index to $p=0$
    and $M[u] \leftarrow -1$ for $0 \le u < \len(\Nt_I)$.
}
\krt{We need a verbal definition of $Q$}
\algstep{S1.}{Input parent loop.}{Set $Q \leftarrow \emptyset$.
}

\algstep{S2.}{Remove ancestry.}{Call Algorithm \algref{R} on each edge $(\ell, r, p, c)$ in $\Et_I$
    corresponding to parent $p$ (which stores the resulting segments in the merge queue $Q$).
}

\algstep{S3.}{Merge ancestors.}{Call Algorithm \algref{M} on $Q$
    (which adds to $\Nt_O$ and $\Et_O$).
}

\algstep{S4.}{Next parent.}{Set $p \leftarrow p+1$;
    if $p = \len(\Nt_I)$, terminate and return $\Nt_O$ and $\Et_O$;
    otherwise, return to \algref{S1}.
}

@@

\begin{taocpalg}{R}{Remove ancestry}
    {Given an edge $(\ell, r, p, c)$,
    and a list $\Al[c]$ of the $m$ ancestral segments corresponding to $c$,
    remove segments overlapping $[\ell, r)$ from $\Al[c]$,
    and add those overlapping segments to $Q$.
    }

    \algstep{R0.}{Initialize.}{
        Set $j \leftarrow 0$.
    }

    \algstep{R1.}{Segment loop.}{
    If $j=m$, terminate; else let $(u, v, x) = \Al[c][j]$.
    (Here $x$ is the output node that inherits from this segment.)
    }

    \algstep{R2.}{Add to merge queue.}{If $[u,v) \cap [\ell,r) = \emptyset$,
    skip to \algref{R4}.
    Otherwise, append $(\max(u,\ell), \min(v,r), x)$ to $Q$.
    }

    \algstep{R3.}{Remove ancestry.}{
    Delete $\Al[c][j]$ and insert in its place
    $(\min(u,\ell), \max(v,\ell), x)$ (if $u<\ell$)
    and(/or) $(\min(v,r), \max(v,r), x)$ (if $r<v$).
    }

    \algstep{R4.}{Iterate.}{Set $j \leftarrow j+1$ and return to \algref{R1}.
    }

@@

\begin{taocpalg}{M}{Merge ancestors}
    {Merge the ancestral segments in $Q$,
    and create the new ancestor $\Al[p]$.
    This algorithm also appends to $\Nt_O$ and $\Et_O$,
    and needs to know about $S$,
    to check if input nodes are samples.
    }

    \algstep{M0.}{Initialize.}{
        Set $z \leftarrow (0, 0, -1)$ and $d = \text{False}$.
    }

    \algstep{M1.}{Segment loop.}{If $\len(Q) = 0$, go to \algref{M9}.}

    \algstep{M2.}{Find next segment overlap boundaries}{
        Let $\ell_0 \leftarrow \min\{x.\lleft : x \in Q\}$
        and $H \leftarrow \{x \in Q : x.\lleft = \ell_0\}$.
        Also let $\ell_+ \leftarrow \min\{x.\lleft : x \in Q \setdiff H\}$
        and $r = \min(\ell_+, \min\{x.\lright : x \in H\})$.
    }

    \algstep{M3.}{Copy single segment.}{
        If $\len(H) = 1$,
        set $\alpha \leftarrow (\ell, r, H[0].\lid)$.
        If also $p \in S$, then
        set $\alpha.\lid \leftarrow M[p]$ and
        $\add(\Et_O, (\alpha.\lleft, \alpha.\lright, \Nt[p], \alpha.\lid))$.
        Go to \algref{M7}.
    }

    \algstep{M4.}{New output node.}{
        If $M[p] = -1$ set $M[p] \leftarrow \add(\Nt_O, \Nt_I[p])$
        Set $\alpha \leftarrow (\ell, r, M[p])$.
    }

    \algstep{M6.}{Record edges and update $Q$.}{
        For each $x \in H$, do $\add(\Et_O, (\ell, r, M[p], x.\lid))$.
    }

    \algstep{M7.}{Update $Q$.}{
        Remove from $Q$ any segment $x \in H$ with $x.\lright = r$
        and then set $x.\lleft = \ell$ for all remaining $x \in H$.
    }

    \algstep{M8.}{Add segment to ancestor.}{
        If $z.\lright = \alpha.\lleft$ and $z.\lid = \alpha.\lid$ set $d = \text{True}$.
        Append $\alpha$ to $\Al[p]$, set $z \leftarrow \alpha$, and return to \algref{M1}.
    }

    \algstep{M9.}{Defragment.}{
        If $d$ is True, find and merge any adjacent segments in $\Al[p]$
        with the same $\lid$.  Terminate.
    }

\end{taocpalg}
\krt{I think the algo discussion is  clear.  I've not tried to implement it, which would be the real test, but I find the description fine other than needing to define $Q$ when first used.}
